An Analysis of Health & Lifestyle Factors Associated with Sleep Quality

### 2024-04-18 products/manuscript/Manuscript.qmd
Added background section, editing of manuscript
Emma - 4/17

For our project, we obtained a dataset from Kaggle.com containing 373
observations of 13 variables from a study on different factors that may
affect sleep quality. Variables of interest from the study included
physical activity level, age, sex, occupation, stress, and several more.
In RStudio (version 4.3), we used a combination of packages to create
models suitable for answering our primary question: which variables are
most important for determining an individual's sleep quality and what
effect do they have? After conducting an exploratory data analysis (EDA)
we hypothesized that stress, sleep duration, and occupation would have
the largest impact on sleep quality. We believed that stress and sleep
quality would be negatively correlated, sleep duration and quality would
be positively correlated, and that jobs with longer or more variable
hours would negatively affect sleep quality.

@@

*Feel free to edit this Emma; I tried to include everything needed*

@@

The data was obtained from kaggle.com at this
[link](https://www.kaggle.com/datasets/henryshan/sleep-health-and-lifestyle/data).
It contains data from a study on individuals of different sexes and age
regarding their sleep quality and different variables that may affect
this. We are unsure of where this data set came from and how it was
collected; the publisher on Kaggle has yet to respond to a comment with
this question posted on March 5th 2024. Regardless, it contains
interesting variables that have been shown in previous studies to have
an effect on sleep quality.

For instance, in Sun et al. (2015), a sample of Chinese individuals
showed a correlation between obesity and worsened sleep quality in men
but not women. We are interested in seeing if this trend holds up for
our dataset; given that most of the observations are of men, we expect
to see a correlation between lower sleep scores and obesity in our data.
Furthermore, it has been shown that age and gender have a notable effect
on sleep quality (Madrid-Valero et al. 2016). Women seem to experience a
deterioration in sleep quality as they age; however, while this trend is
still present in men, it is notably less consistent and can vary
dramatically between individuals. We expect to see similar trends in our
dataset, but as stated previously, our observations contain
significantly more men than women. Finally, we wish to test whether or
not certain occupational industries have different effects on sleep
quality. It has been shown that managerial positions tend to have the
worst sleep quality among civilian sector workers while 24 jobs that
have rotating shifts tend to have the worst sleep quality (Luckhaupt et
al. 2010). Our dataset contains observations of mostly
white-collar/service-based jobs, so we expect to find similar trends
upon analysis.

*NEED TO INSERT*

*EMMA'S SOURCES ARE LINKED HERE:*

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8345503/ --\> blood
pressure and sleep
https://www.apa.org/news/press/releases/stress/2013/sleep --\> sleep
duration & stress on sleep quality
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC10503965/#:\~:text=Regular%20physical%20activity%20can%20lead,managing%20sleep%20disorders%20like%20insomnia.
--\> physical activity and sleep quality

@@

The dataset originally contained 373 observations for 13 variables
covering a broad spectrum health and lifestyle variables associated with
sleep quality. The curators of the dataset did not provide information
on how or where this data was collected, so we solely relied on the
codebook for variable definitions. There were no missing values or
erroneous variables in the dataset that we needed to remove. The
"Blood.Pressure" variable was transformed to reflect the systolic and
diastolic blood pressure measurements from each subject. We then chose
to add an additional variable, labeled 'cat_bp' to reflect the
categorical blood pressure status of a subject based on the [*American
Heart Association
guidelines*](https://www.heart.org/en/health-topics/high-blood-pressure).
Additionally we created a variable, 'StepsGroup' to determine the
categorical activity level of subjects in the dataset based on these
guidelines provided by 10,000 steps [*10,000
steps.org*](https://www.10000steps.org.au/articles/healthy-lifestyles/counting-steps/#:~:text=The%20following%20pedometer%20indices%20have%20been%20developed%20to,day%205%20Highly%20active%20is%20more%20than%2012%2C500)*.*
Finally, we created an additional categorical variable,
'PhysicalActivityGroup' based on 30 minute increments, to categorically
represent the level of physical activity reported by a subject. The
outcome of interest is noted as "Quality.of.Sleep" that we will
reference as Sleep Quality or Quality of Sleep throughout this report.

@@

This analysis was conducted under R version 4.3.1 on a MacOS operating
system *cite R*. The following R packages were used in the development
of these analyses: *R PACKAGES USED: here, skimr, broom, tidyverse,
ggplot2, dplyr, corrplot, ranger, vip* All processing and analysis code
can be found in the Supplementary Material file.

@@

*To Emma: did you want this text in both places you placed it?* A
variable importance plot (VIP) was constructed to determine which
variables had the strongest impact in our Random Forest Model. *INCLUDE
PLOT HERE* Sleep Duration, Stress Level, and Age had the highest levels
of importance in our model. Physical Activity Level was excluded
entirely, which was to be expected given our exploratory data analysis
and simple linear regression models.

@@

>>>>

For our project, we obtained a dataset from [Kaggle](https://www.kaggle.com/datasets) containing 373 observations of 13 variables from a study on different factors that may
affect sleep quality. Variables of interest from the study included
physical activity level, age, sex, occupation, stress, gender, BMI, blood pressure, and sleep duration.
In [RStudio](https://posit.co/download/rstudio-desktop/) (version 4.3.1), we used a combination of packages to create
models suitable for answering our primary question: which variables are
most important for determining an individual's sleep quality and what
effect do they have? After conducting an exploratory data analysis (EDA)
we hypothesized that stress, sleep duration, and occupation would have
the largest impact on sleep quality. We believed that stress and sleep
quality would be negatively correlated, sleep duration and quality would
be positively correlated, and that jobs with longer or more variable
hours would negatively affect sleep quality.

@@

@@

The data was obtained from kaggle.com at this
[link](https://www.kaggle.com/datasets/henryshan/sleep-health-and-lifestyle/data).
The data contains observations from a study on individuals of varying sexes and age
regarding their sleep quality and different variables that may affect
this. We are unsure of where this data set came from and how the data were
collected; the publisher on Kaggle has yet to respond to a comment with
this question posted on March 5th 2024. Regardless, it contains
interesting variables that have been shown in previous studies to have
an effect on sleep quality.

For instance, in Sun et al. (2015), a sample of Chinese individuals
showed a correlation between obesity and worsened sleep quality in men
but not women. We are interested in seeing if this trend in present in 
our dataset; however, given that most of the observations are of men, we expect
to see a correlation between lower sleep scores and obesity in our data.
Furthermore, it has been shown that age and gender have a notable effect
on sleep quality (Madrid-Valero et al. 2016). Women seem to experience a
deterioration in sleep quality as they age; however, while this trend is
still present in men, it is notably less consistent and can vary
dramatically between individuals. We expect to see similar trends in our
dataset, but as stated previously, our observations contain
significantly more men than women. Finally, we wish to test whether or
not certain occupational industries have different effects on sleep
quality. It has been shown that managerial positions tend to have the
worst sleep quality among civilian sector workers while 24 jobs that
have rotating shifts tend to have the worst sleep quality (Luckhaupt et
al. 2010). Our dataset contains observations of mostly
white-collar/service-based jobs, so we expect to find similar trends
upon analysis.

Additionally, based on the findings of Sun et. al, that obesity can negatively impact sleep quality, we also expect that higher levels of blood pressure will result in poorer sleep quality. The American Sleep Foundation states that both short and long sleep duration increase risk of hypertension in adults (Li & Shang, 2021). This draws interesting focus to the circular nature of sleep quality and hypertension. Those with hypertension also face a higher risk of insomnia, a disorder affecting the ability to fall asleep, maintain sleep, and sleep without continuous disturbance (Liu et. al, 2022). Short sleep duration is also positively associated with stress level (APA, 2013). Survey findings indicate that the average American sleeps 6.7 hours per night, which is well below the minimum recommended duration seven to nine hours (APA, 2013). Similar to the relationship between hypertension and sleep quality, as stress level increases, sleep duration decreases thereby causing poorer sleep quality. Poorer sleep quality then leads to higher stress, greater risk of hypertension, and shorter sleep duration.

Engaging in physical activity has been suggested to improve both sleep quality and sleep duration. Regular moderate to intense physical activity stimulates an increase of melatonin production, a hormone that regulates sleep-wake cycles (Alnawwar et. al, 2023). Additionally, moderate to intense exercise has shown to have a positive effect on those with insomnia (Alnawwar et. al, 2023). Physical activity also reduces stress and improves both sleep quality and sleep efficiency (Alnawwar et. al, 2023). However, studies also show that frequent high-intensity workouts could lead to poorer sleep quality among people with and without insomnia (Alnawwar et. al, 2023). Studies indicate that consistent moderate exercise three to four times a week (approximately 150 minutes total) is the optimal weekly regimen to improve sleep quality and reduce stress. We expect to see similar trends with our data, where increased stress levels negatively impact sleep quality and sleep duration, while noticing a positive association between increased physical activity levels and sleep quality.

@@

The dataset originally contained 373 observations for 13 variables
covering a broad spectrum health and lifestyle variables associated with
sleep quality. The curators of the dataset did not provide information
on how or where this data was collected, so we solely relied on the
codebook for variable definitions. There were no missing values or
erroneous variables in the dataset that we needed to remove. The
"Blood.Pressure" variable was transformed to reflect the individual systolic and
diastolic blood pressure measurements from each subject. We then chose
to add an additional variable, labeled 'cat_bp' to reflect the
categorical blood pressure status of a subject based on the [American
Heart Association
guidelines](https://www.heart.org/en/health-topics/high-blood-pressure).
Additionally we created a variable, 'StepsGroup' to determine the
categorical activity level of subjects in the dataset based on these
guidelines provided by [10,000 steps.org] (https://www.10000steps.org.au/articles/healthy-lifestyles/counting-steps/#:~:text=The%20following%20pedometer%20indices%20have%20been%20developed%20to,day%205%20Highly%20active%20is%20more%20than%2012%2C500)
Finally, we created an additional categorical variable,
'PhysicalActivityGroup' based on 30 minute increments, to categorically
represent the level of physical activity reported by a subject. The
outcome of interest is noted as "Quality.of.Sleep" that we will
reference as Sleep Quality or Quality of Sleep throughout this report.

@@

This analysis was conducted under R version 4.3.1 on a MacOS operating
system. The following R packages were used in the development
of these analyses:here, skimr, broom, tidyverse,
ggplot2, dplyr, corrplot, ranger, and vip. All processing and analysis code
can be found in the Supplementary Material file.

@@

*To Emma: did you want this text in both places you placed it?*
*To Kevin: I placed in both sections as a placeholder, whichever you think is best!*

A
variable importance plot (VIP) was constructed to determine which
variables had the strongest impact in our Random Forest Model. *INCLUDE
PLOT HERE* Sleep Duration, Stress Level, and Age had the highest levels
of importance in our model. Physical Activity Level was excluded
entirely, which was to be expected given our exploratory data analysis
and simple linear regression models.

@@

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8345503/ --\> blood
pressure and sleep

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9815923/#:~:text=Patients%20with%20hypertension%20commonly%20complain,an%20increased%20risk%20of%20insomnia. --\> hypertension and insomnia

https://www.apa.org/news/press/releases/stress/2013/sleep --\> sleep
duration & stress on sleep quality

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC10503965/#:\~:text=Regular%20physical%20activity%20can%20lead,managing%20sleep%20disorders%20like%20insomnia.
--\> physical activity and sleep quality

https://www.heart.org/en/health-topics/high-blood-pressure

https://www.10000steps.org.au/articles/healthy-lifestyles/counting-steps/#:~:text=The%20following%20pedometer%20indices%20have%20been%20developed%20to,day%205%20Highly%20active%20is%20more%20than%2012%2C500
